Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-08 Manuscript/manuscript.tex
Finished writing observations and results to Stoichiometry section -- still has rugged style but the hard data shall be all there.

Binding stoichiometry of \ce{Ca2+} and POPC was 
thoroughly studied in the experimental work~\citenum{altenbach84}.
Several binding models were proposed and tested of which only one,
ternary complex binding model, provided a good fit of the experimental observations. 
In such a model, it is assumed that \ce{Ca2+} cations bind to a POPC membrane 
with a stoichiometry 2~POPC:1~\ce{Ca^{2+}}.
Such a model also provides a good fit 
to our simulations with ECC-lipids 
(see Fig.~\ref{fig:cacl-bind} in SI and its caption for details). 
Simulations, however, allow us to directly evaluate the stoichiometry 
by calculating relative propensities of various \ce{Ca2+}:$n\times$POPC clusters
In Figure~\ref{fig:cacl_complexes} we see that ternary complex is indeed 
the most probable binding mode of calcium at 285~mM concentration. 
Apart from this complex, we also find complexes with 1~and 3~lipids
occuring with only a slightly lower but similar probability.
The fractions of \ce{Ca2+}:$n\times$POPC complexes at 285~mM concetration 
are then in order:
$42\%$ for two lipids,
$30\%$ for one lipid, and
$28\%$ for three lipids.
The symmetry of such a distribution --  
i.e. almost equal probabilities of complexes with 1 or 3 lipids 
that behave in the total average picture as complexes with 2 lipids -- 
provides clues why ternary complex binding model 
fits both simulation and experimental results relatively well, 
although it is apparently incorrect.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/number_bound_lipids_p_cation_Ecc-lipids.eps} \\
  \caption{\label{fig:cacl_complexes}
      Histogram of relative probabilities of existence of complexes
      of \ce{Ca2+} and $n$ POPC lipids. 
  }
\end{figure}

In addition, we estimated the relative binding affinities of several moieties in POPC:
phosphate moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
Using the probability isodensity contours (see Fig.~\ref{fig:volmaps}), 
we estimate that the largest contribution
to the binding of \ce{Ca2+} to POPC membranes likely comes from the phosphate group. 
Although the isodensity plots are relatively easy to interpret,
the contours shown in Fig.~\ref{fig:volmaps} cannot conclusively tell on the details
of \ce{Ca2+} binding to any of the two carbonyl moieties.

\todo{JOE: Put details of the cation-binding stoichiometry analysis to SI.} \\
\todo{JOE: Update the binding isotherm figure with new simulations} \\
\todo{SAMULI: The same authors have also literature, where they say that ternary complex may not
  be the only option. I will recheck and come back to this.
  SAMULI: This is written in \cite{macdonald87}:
  ''Ca2+binding to POPC bilayers over the whole concentration range can be best described in terms
  of formation of a ternary complex involving complexation of two lipids to one calcium ion (Altenbach and Seelig, 1984).
  The addition of a sodium competition term has not changed this conclusion. However, if Ca2+ concentrations
  up to 100 mM are considered, the data can be equally well explained by a 1:1 binding mechanism (cf. Figure 7).
  In contrast, the Ca2+ binding to POPC-POPG mixtures can be best described by assuming a 1:l stoichiometry regardless of the range of Ca2+concentrations.''
  We might or might not want to discuss about this.} \\
\todo{SAMULI: I would also analyze how much there is contact between ions and different
parts of the lipid (phosphase, carbonyl, etc.).
JOE: Pavel suggest using volumetric map (e.g. in VMD) to visualise the amounts of Ca, Na and Cl bound to different moieties.} \\
\todo{SAMULI: I think we should quantify this, i.e. how much there are these.
  Maybe also the other possible complexes? Maybe also the correlation betweem complexes
  and binding sites, if it is not too much work.
  JOE: This looks like a careful work for the next paper to me. 
  I'd only add a relatively simple analysis of binding sites and probably the propensity of 1-2-3 membered clusters.
  SAMULI: I agree that we should not use too much time on this now.}. \\
\todo{Finalize stoichiometry analysis for \ce{Na^+}, \ce{Ca^{2+}},
  their interaction energies with the lipid membrane, etc,
  and finalize the discussion after these results.}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/volmap_resid10_Ca_Cl_PO4Cent.png} \\
  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
      Contours of probability isodensities of \ce{Ca2+} with respect to 
      various moieties fixed in space (highlighted with transparent spheres): 
      phosphate moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
      Shown contours suggest that the dominant contribution 
      to \ce{Ca2+} binding likely comes from the phosphate group, 
      but remain inconclusive on the details of the interactions with any of the two carbonyl groups. 
  }
  \todo{JOE: I'll update this figure with some ensemble of configuration to support binding preference of \ce{Ca2+}}
\end{figure}

%\begin{figure}[tbp]
%  \centering
%  \includegraphics[height=6.0cm]{../Fig/POPC_binding_labels.eps} 
%  \includegraphics[height=6.0cm]{../Fig/POPC_binding_ECC-lipids.eps}
%  \caption{\label{fig:binding_conf}
%      Configurations of \ce{Ca2+} binding in POPC membranes in the ECC-lipids model.
%      The binding sites are: 
%      red circle for phosphate oxygens; 
%      blue square for sn-1 carbony oxygen; and green triangle for sn-2 carbonyl oxygen.
%      The configurations along with their probabilities are drawn in the following manner:
%      each row corresponds to an individual lipid the cation is bound to;
%      symbols denote the site the cations is bound to the respective lipid.
%      An empty configuration denotes an unbound cation in a solvent. 
%  }
%\end{figure}

@@

It was found in the original work~\cite{altenbach84} that 
a ternary complex binding model (i.e.~2~POPC:1~\ce{Ca^{2+}})
provides the best fit to experimental measurements of all considered models in that study. 
In such a model, there is a linear relationship between quantities 
$C_b$, mole fraction of bound \ce{Ca2+} per POPC, and $\sqrt{C_b/C_I}$, 
where $C_I$ is the concentration of free cations at the plane of ion binding~\cite{altenbach84}.
The concentration $C_b$ was obtained from an extrapolation of linear relation 
between deuterium NMR measurements and atomic absorption spectroscopy for low concetrations of \ce{CaCl2}.
Such an extrapolation is valid as long as the mode of \ce{Ca2+} binding 
remains constant throughout the extrapolation range. 
The concentration $C_I$ is determined by using the surface potential by using the Boltzmann equation.
However, Boltzmann theory yields inaccurate results
for divalent cations like \ce{Ca2+}~\cite{Andelmann1995}. 
An atomistic simulation, on the other hand, provides these quantities directly without severe assumptions. 
Hence we hypothesise that the discrepancy between the results in the experiment~\cite{altenbach84} and 
our simulations likely lays in the fact that 
the assumptions and relations used for determinig concentrations $C_b$ and $C_I$ in the experiment~\cite{altenbach84}
gradually do not hold for higher concentrations of \ce{Ca2+}.

\begin{figure}[]
  \centering
  \includegraphics[height=9.0cm,angle=-90]{../Fig/bound-CAs_conc-eccl17.eps}
  \caption{\label{fig:cacl-bind}
    Ternary complex binding model of \ce{Ca2+} to a POPC membrane 
    that assumes the stoichiometry of 2~POPC:1~\ce{Ca^{2+}} (details in reference~\citenum{altenbach84}) 
    provides a good fit to experimental measurements~\cite{altenbach84}
    and it also provides a good fit to our simulation data. 
    Note that the units in the reference~\citenum{altenbach84} are different from the units presented here,
    and, hence, the observed slope of the linear relationship is different.
    }
\end{figure}

>>>>

Binding stoichiometry of \ce{Ca^{2+}} and POPC was 
thoroughly studied in the experimental work~\cite{altenbach84}.
Several binding models were proposed and tested of which only one,
ternary complex binding model, provided a good fit of the experimental observations. 
In such a model, it is assumed that \ce{Ca^{2+}} cations bind to a POPC membrane 
with a stoichiometry 2~POPC:1~\ce{Ca^{2+}}.
In a later work~\cite{macdonald87}, 
a Langmuir adsorption model (i.e. stoichiometry 1~POPC:1~\ce{Ca^{2+}}) 
was found to provide as good fit as ternary complex model, 
when only low concentrations of \ce{CaCl2} are considered.

Ternary complex model also provides a good fit 
to our simulations with ECC-lipids 
(see Fig.~\ref{fig:cacl-bind} in SI and its caption for details). 
Simulations, however, allow us to directly evaluate the stoichiometry 
by calculating relative propensities of various \ce{Ca^{2+}}:$n\times$POPC clusters
In Figure~\ref{fig:cacl_complexes} we see that ternary complex is indeed 
the most probable binding mode of calcium at 285~mM concentration. 
Apart from this complex, we also find complexes with 1~and 3~lipids
occuring with only a slightly lower but similar probability.
The fractions of \ce{Ca^{2+}}:$n\times$POPC complexes at 285~mM concetration 
are then in order:
$42\%$ for two lipids,
$30\%$ for one lipid, and
$28\%$ for three lipids.
The symmetry of such a distribution --  
i.e. almost equal probabilities of complexes with 1 or 3 lipids 
that behave in the total average picture as complexes with 2 lipids -- 
provides clues why ternary complex binding model 
fits both simulation and experimental results relatively well, 
although it is apparently incorrect.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/number_bound_lipids_p_cation_Ecc-lipids.eps} \\
  \caption{\label{fig:cacl_complexes}
      Histogram of relative probabilities of existence of complexes
      of \ce{Ca^{2+}} and $n$ POPC lipids. 
  }
\end{figure}

In addition, we estimated relative binding affinities 
of several moieties in POPC towards \ce{Ca^{2+}}. 
Based on the probability isodensity contours (see Fig.~\ref{fig:volmaps}), 
we estimate that the largest contribution
to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
Although the isodensity plots are relatively easy to interpret,
the contours shown in Fig.~\ref{fig:volmaps} cannot conclusively tell on the details
of \ce{Ca^{2+}} binding to any of the two carbonyl moieties
but their apparently lower affinity compared to phosphate oxygens.

The residence time of \ce{Ca^{2+}} bound to a POPC membrane 
is experimentally determined to be lower than $10\,\mu\mathrm{s}$~\cite{carrington67}. 
Recent theoretical work with long enough simulations estimates this time
in the order of $1\,\mu\mathrm{s}$~\cite{catte16}. 
This is in contrast to our model, ECC-lipids, 
which gives a mean residence time in the order of $10\,\mathrm{ns}$, 
i.e. at least two orders of magnitude lower than previous estimates.
Such a finding changes the point of view of calcium binding from
very tight long-term stable binding with rare exchanges to 
a relatively frequent exchange of cations 
in equilibrium between membrane and solvent.

\todo{SAMULI: The same authors have also literature, where they say that ternary complex may not
  be the only option. I will recheck and come back to this.
  SAMULI: This is written in \cite{macdonald87}:
  ''\ce{Ca^{2+}} binding to POPC bilayers over the whole concentration range can be best described in terms
  of formation of a ternary complex involving complexation of two lipids to one calcium ion (Altenbach and Seelig, 1984).
  The addition of a sodium competition term has not changed this conclusion. However, if \ce{Ca^{2+}} concentrations
  up to 100 mM are considered, the data can be equally well explained by a 1:1 binding mechanism (cf. Figure 7).
  In contrast, the Ca2+ binding to POPC-POPG mixtures can be best described by assuming a 1:l stoichiometry regardless of the range of Ca2+concentrations.''
  We might or might not want to discuss about this.
  JOE: I think this is adressed/acknowledged enough now.} \\
\todo{Finalize stoichiometry analysis for \ce{Na^+}, \ce{Ca^{2+}},
  their interaction energies with the lipid membrane, etc,
  and finalize the discussion after these results.}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/volmap_resid10_Ca_Cl_PO4Cent.png} \\
  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
      Contours of probability isodensities of \ce{Ca^{2+}} with respect to 
      various moieties fixed in space (highlighted with transparent spheres): 
      phosphate moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
      Shown contours suggest that the dominant contribution 
      to \ce{Ca^{2+}} binding comes from the phosphate oxygens,
      whereas the interactions with any of the two carbonyl groups are considerably milder.
  }
  \todo{JOE: I'll update this figure with some ensemble of configuration to support binding preference of \ce{Ca^{2+}}}
\end{figure}

%\begin{figure}[tbp]
%  \centering
%  \includegraphics[height=6.0cm]{../Fig/POPC_binding_labels.eps} 
%  \includegraphics[height=6.0cm]{../Fig/POPC_binding_ECC-lipids.eps}
%  \caption{\label{fig:binding_conf}
%      Configurations of \ce{Ca^{2+}} binding in POPC membranes in the ECC-lipids model.
%      The binding sites are: 
%      red circle for phosphate oxygens; 
%      blue square for sn-1 carbony oxygen; and green triangle for sn-2 carbonyl oxygen.
%      The configurations along with their probabilities are drawn in the following manner:
%      each row corresponds to an individual lipid the cation is bound to;
%      symbols denote the site the cations is bound to the respective lipid.
%      An empty configuration denotes an unbound cation in a solvent. 
%  }
%\end{figure}

@@

It was found in the original work~\cite{altenbach84} that 
a ternary complex binding model (i.e.~2~POPC:1~\ce{Ca^{2+}})
provides the best fit to experimental measurements of all considered models in that study. 
In such a model, there is a linear relationship between quantities 
$C_b$, mole fraction of bound \ce{Ca^{2+}} per POPC, and $\sqrt{C_b/C_I}$, 
where $C_I$ is the concentration of free cations at the plane of ion binding~\cite{altenbach84}.
The concentration $C_b$ was obtained from an extrapolation of linear relation 
between deuterium NMR measurements and atomic absorption spectroscopy for low concetrations of \ce{CaCl2}.
Such an extrapolation is valid as long as the mode of \ce{Ca^{2+}} binding 
remains constant throughout the extrapolation range. 
The concentration $C_I$ is determined by using the surface potential by using the Boltzmann equation.
However, Boltzmann theory yields inaccurate results
for divalent cations like \ce{Ca^{2+}}~\cite{Andelman1995}. 
An atomistic simulation, on the other hand, provides these quantities directly without severe assumptions. 
Hence we hypothesise that the discrepancy between the results in the experiment~\cite{altenbach84} and 
our simulations likely lays in the fact that 
the assumptions and relations used for determinig concentrations $C_b$ and $C_I$ in the experiment~\cite{altenbach84}
gradually do not hold for higher concentrations of \ce{Ca^{2+}}.

\begin{figure}[]
  \centering
  \includegraphics[height=9.0cm,angle=-90]{../Fig/bound-CAs_conc-eccl17.eps}
  \caption{\label{fig:cacl-bind}
    Ternary complex binding model of \ce{Ca^{2+}} to a POPC membrane 
    that assumes the stoichiometry of 2~POPC:1~\ce{Ca^{2+}} (details in reference~\citenum{altenbach84}) 
    provides a good fit to experimental measurements~\cite{altenbach84}
    and it also provides a good fit to our simulation data. 
    Note that the units in the reference~\citenum{altenbach84} are different from the units presented here,
    and, hence, the observed slope of the linear relationship is different.
    }
\end{figure}
